---
title: "Computing in R"
subtitle: "Amsterdam UMC Doctoral School"
author: "Perry Moerland"
date: "14-18 March 2021"
output:
  xaringan::moon_reader:
    self_contained: TRUE
    nature:
      beforeInit: "macros.js"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE,echo=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(prompt=TRUE, comment='#',fig.path='Figures/')
```

# Graphics

R has versatile tools for graphics. There are typically three steps to
producing useful graphics:

1. Creating the basic plot

1. Enhancing the plot with labels, legends, colors etc.

1. Exporting the plot from R for use elsewhere
  
---
# Basic plot (I)

It is straightforward to make a simple plot using functions from the
__graphics__ package (loaded by default):

```{r poly1, fig.height=3.5, dev='svg',echo=c(2,3)}
par(mar = c(4, 4, .1, .1))
x <- (0:100)/10
plot(x, x^3 - 13 * x^2 + 39 * x)
```
---
# Basic plot (II)
You can increase the size of the symbols on the axes and the axis labels (_cex_ stands for character expansion factor):

```{r poly2, fig.height=3.5, dev='svg',echo=2}
par(mar = c(4, 5, .1, .1))
plot(x, x^3 - 13 * x^2 + 39 * x,cex.axis=1.3,cex.lab=1.5)
```

---
# Basic plot (III)

Change the type of plot via the argument _type_: "p" for points (is default), "l" for lines, etc. See [?plot](https://www.rdocumentation.org/packages/graphics/versions/3.5.1/topics/plot) for other options

```{r poly3, fig.height=3.5, dev='svg',echo=2}
par(mar = c(4, 5, .1, .1))
plot(x,x^3-13*x^2+39*x,type="l",cex.axis=1.3,cex.lab=1.5)
```

---
# Basic plot (IV)

- Change the titles for the axes via _xlab_ and _ylab_
- Add an overall title for the plot via _main_

```{r poly4, fig.height=3.5, dev='svg',echo=2}
par(mar = c(4, 5, 2, .1))
plot(x,x^3-13*x^2+39*x,type="l",xlab="time (hours)",
     ylab="temperature",main="Enhanced plot",cex.axis=1.3,
     cex.lab=1.5)
```
---
# Basic plot (V)
- Change the plot symbol used from the default dot using the argument _pch_ (see [reference card](http://bioinformaticslaboratory.eu/wp-content/uploads/gs-computing-in-r/R-refcard-v2.pdf))
- Change the colour via the argument _col_. By name: see `colors()` for the 657 [options](https://rstudio-pubs-static.s3.amazonaws.com/3486_79191ad32cf74955b4502b8530aad627.html). By number: see `palette()`

```{r poly5, fig.height=3, dev='svg',echo=2}
par(mar = c(4, 5, 2, .1))
plot(x,x^3-13*x^2+39*x,pch=18,xlab="time (hours)",
     ylab="temperature",col="red",main="Enhanced plot",
     cex.axis=1.3,cex.lab=1.5)
```

---
# Plot symbols

There are 25 different plot symbols, see [?points](https://www.rdocumentation.org/packages/graphics/versions/3.5.1/topics/points)

```{r pch, fig.height=4, dev='svg', fig.show='hold'}
plot(1:25, pch=1:25,cex=2,bg="grey")
# bg: background colors for open plot symbols
```

---
# Enhancing a plot

.pull-left[
```{r stepwise-plot1,fig.show='hide',prompt=FALSE}
x<-(0:100)/10
plot(x,x^3-13*x^2+39*x,
     type="l",
     xlab="time (hours)",
     ylab="temperature",
     cex.axis=1.5,cex.lab=1.5)
```
]

.pull-right[
![](`r knitr::fig_chunk("stepwise-plot1", "png")`)
]

---
# Enhancing a plot

.pull-left[
```{r stepwise-plot2,fig.show='hide',prompt=FALSE}
x<-(0:100)/10
plot(x,x^3-13*x^2+39*x,
     type="l",
     xlab="time (hours)",
     ylab="temperature",
     cex.axis=1.5,cex.lab=1.5)
points(2,34,col="red",pch=16,
         cex=2)
```
]

.pull-right[
![](`r knitr::fig_chunk("stepwise-plot2", "png")`)
]
---
# Enhancing a plot

.pull-left[
```{r stepwise-plot3,fig.show='hide',prompt=FALSE}
x<-(0:100)/10
plot(x,x^3-13*x^2+39*x,
     type="l",
     xlab="time (hours)",
     ylab="temperature",
     cex.axis=1.5,cex.lab=1.5)
points(2,34,col="red",pch=16,
       cex=2)
{{arrows(4,50,2.2,34.5)}}
```
]

.pull-right[
![](`r knitr::fig_chunk("stepwise-plot3", "png")`)
]
---
# Enhancing a plot

.pull-left[
```{r stepwise-plot4,fig.show='hide',prompt=FALSE}
x<-(0:100)/10
plot(x,x^3-13*x^2+39*x,
     type="l",
     xlab="time (hours)",
     ylab="temperature",
     cex.axis=1.5,cex.lab=1.5)
points(2,34,col="red",pch=16,
       cex=2)
arrows(4,50,2.2,34.5)
text(4.15,50,"local maximum",
     adj=0,col="blue",cex=1.5)
```
]

.pull-right[
![](`r knitr::fig_chunk("stepwise-plot4", "png")`)
]

---
# Enhancing a plot

.pull-left[
```{r stepwise-plot5,fig.show='hide',prompt=FALSE}
x<-(0:100)/10
plot(x,x^3-13*x^2+39*x,
     type="l",
     xlab="time (hours)",
     ylab="temperature",
     cex.axis=1.5,cex.lab=1.5)
points(2,34,col="red",pch=16,
       cex=2)
arrows(4,50,2.2,34.5)
text(4.15,50,"local maximum",
     adj=0,col="blue",cex=1.5)
lines(x,30-50*sin(x/2),
      col="blue")
```
]

.pull-right[
![](`r knitr::fig_chunk("stepwise-plot5", "png")`)
]

---
# Enhancing a plot

.pull-left[
```{r stepwise-plot6,fig.show='hide',prompt=FALSE}
x<-(0:100)/10
plot(x,x^3-13*x^2+39*x,
     type="l",
     xlab="time (hours)",
     ylab="temperature",
     cex.axis=1.5,cex.lab=1.5)
points(2,34,col="red",pch=16,
       cex=2)
arrows(4,50,2.2,34.5)
text(4.15,50,"local maximum",
     adj=0,col="blue",cex=1.5)
lines(x,30-50*sin(x/2),
      col="blue")
legend(x=0,y=80,
       legend=c("Sahara",
                "Gobi"),
       lty=1,
       col=c("black","blue"),
       cex=1.5)
```
]

.pull-right[
![](`r knitr::fig_chunk("stepwise-plot6", "png")`)
]

---
# Graphical parameters (I)

You can change the default value of many graphical parameters via _par_ (see [?par](https://www.rdocumentation.org/packages/graphics/versions/3.5.1/topics/par)). For example to reset the background of a plot to
green:
```{r par1}
par("bg")
```

--

```{r par2}
par(bg="green")
par("bg")
```

---
# Graphical parameters (II)
Any plot that you'll make from now on will have a green background:

```{r par3, fig.height=4, dev='svg', message=FALSE, warning=FALSE,echo=2}
par(bg="green")
plot(x, x^3 - 13 * x^2 + 39 * x)
```

---
# Graphical parameters (III)

But luckily you can switch back:

```{r par4, fig.height=4, dev='svg', message=FALSE, warning=FALSE}
par(bg="white")
plot(x, x^3 - 13 * x^2 + 39 * x)
```

---
# Graphical parameters: saving defaults

```{r par5, fig.height=4, dev='svg', message=FALSE, warning=FALSE}
par.defaults <- par(no.readonly=TRUE)
par(bg="green")
par(par.defaults)
plot(x, x^3 - 13 * x^2 + 39 * x)
```
---
# Graphical parameters (IV)

Other often used parameters/arguments:
- _lwd_ sets the line width
- _las_ to rotate axis symbols
- _mfrow_ and _mfcol_ enable multiple plots in one figure
- _mar_ to change the default margins of the figure

```{r par6, fig.height=3, dev='svg', message=FALSE, warning=FALSE,echo=2}
par(mar = c(4, 5, .1, .1))
plot(x,x^3-13*x^2+39*x,type="l",xlab= "time (hours)",
     ylab="temperature",lwd=3,las=1,cex.axis=1.3,cex.lab=1.5)
```

---
# Histogram

Use `hist` for plotting histograms. See [?hist](https://www.rdocumentation.org/packages/graphics/versions/3.5.1/topics/hist) for the many arguments of this function.

```{r hist,fig.height=4,echo=3}
library(foreign)
titanic3 <- read.dta("../Exercises/titanic3.dta")
hist(titanic3$age,breaks=15,freq=FALSE,cex.axis=1.5,cex.lab=1.5)
```

---
# Boxplot (I)

.pull-left[
The function `boxplot` can be used on a vector:
```{r boxplot1,fig.show='hide',fig.height=5,prompt=FALSE}
boxplot(titanic3$fare,
        ylim=c(0,300),
        ylab="fare", 
        cex.axis=1.5,
        cex.lab=1.5
        )
```
]
.pull-right[
![](`r knitr::fig_chunk("boxplot1", "png")`)
]

---
# Boxplot (II)

.pull-left[
Or using the formula interface (with ~):
```{r boxplot2,fig.show='hide',fig.height=5,prompt=FALSE}
boxplot(fare~pclass,
        data=titanic3,
        ylim=c(0,300),
        ylab="fare", 
        cex.axis=1.5,
        cex.lab=1.5
        )
```
]
.pull-right[
![](`r knitr::fig_chunk("boxplot2", "png")`)
]


---
# Interactive plots

```{r plotly, fig.height=4, dev='svg', message=FALSE, warning=FALSE}
library(plotly)
dat = data.frame(x=(0:100)/10,y=x^3-13*x^2+39*x)
plot_ly(data=dat,x=~x,y=~y)
```

---
class: center, middle

# Thanks!

Slides created via the R package [**xaringan**](https://github.com/yihui/xaringan).
